\section{Haskell Chase Implementation}

	The goal of the implementation of the chase is to deterministically find
	all possible outcomes of the chase. It does this by forking and taking all
	paths when encountering a disjunct rather than nondeterministically
	choosing one disjunct to satisfy.

	The results from the attempts to satisfy each disjunct are returned as a
	list. The returned list will not contain an entry for runs that return no
	model, and will merge lists returned from runs that themselves encountered
	a disjunct. The lazy evaluation of Haskell allows a user to access members
	of the returned list even though some chase runs have not returned a value.

	Appendix B contains the chase-running portions of the implementation.

	\subsection{Operation}

	The first step of the chase implementation is to make sure that each
	formula of the given theory can be represented as a geometric logic
	formula. If a formula $\varphi$ can not be coerced to a geometric logic
	formula, the chase tries to coerce it into one by applying following rules
	recursively:

	\begin{itemize}
	\item $\neg\alpha \wedge \neg\beta \mapsto \alpha \vee \beta$
	\item $\neg\alpha \vee \neg\beta \mapsto \alpha \wedge \beta$
	\item $\neg\neg\alpha \mapsto \alpha$
	\item $\neg\alpha \to \beta \mapsto \alpha \vee \beta$
	\item $\alpha \to \neg\beta \mapsto \alpha \wedge \beta$
	\item $\neg(\forall\ \vec{x} : \neg\alpha) \mapsto \exists \vec{x} : \alpha$
	\end{itemize}

	All other constructs are preserved. If this transformed formula is still
	not in positive existential form, an error is thrown.

	The chase function then sorts the input formul{\ae} by the number of
	disjunctions on the right side of the implication. It allows branches to
	terminate without growing to an unnecessarily (and possibly infinitely)
	large size. This step will cause each branch of the algorithm to finish in
	less time, as they are likely to halt before branching yet again. The
	formul{\ae} are not sorted purely by absolute number of disjunctions on the
	right side, but by whether there are zero, one, or many disjunctions. This
	is done to avoid unnecessary re-ordering for no gain because formul{\ae}
	with no disjunctions or only a single disjunct on the right side are more
	likely to cause a branch to stop growing than one with many disjunctions.
	Likewise, formul{\ae} with zero disjunctions are more likely to cause a
	branch to halt than those with one or more disjunctions. A secondary sort
	also occurs within these classifications that orders formul{\ae} by the
	number of variables to reduce the number of bindings generated.

	Once the input formul{\ae} are sorted, the $chase$ function begins
	processing a \emph{pending} list, which is initially populated with a
	single model that has an empty domain and no facts.

	For each \emph{pending} model, each formula is evaluated to see if it holds
	in the model for all environments. If an envionment is found that does not
	satisfy the model, the model and environment in which the formula did not
	hold is passed to the $satisfy$ function, along with the formula that needs
	to be satisfied. The list of models returned from $satisfy$ is merged into
	the \emph{pending} list, and the result of running $chase$ on the new
	\emph{pending} list is returned. If, however, the model holds for all
	formul{\ae} in the theory and all possible associated environments, it is
	concatenated with the result of running the chase on the rest of the models
	in the \emph{pending} list and returned.

	The $satisfy$ function performs a pattern match on the type of formula
	given. Assuming $satisfy$ is given a model $\mathbb{M}$, an environment
	$\lambda$, and a formula $\varphi$, $satisfy$ will behave as outlined
	in the following algorithm.

	~\\

	\begin{algorithm}[H]
	\DontPrintSemicolon
	\TitleOfAlgo{satisfy :: Model $\to$ Environment $\to$ Formula $\to$ [Model]}
	return \Switch{$\varphi$}{
		\lCase{$\top$}{return a list containing $\mathbb{M}$} \;
		\lCase{$\bot$}{return an empty list} \;
		\lCase{$x=y$}{return a list containing $quotient(\mathbb{M})$} \;
		\lCase{$\alpha \vee \beta$}{return $satisfy(\alpha) \cup satisfy(\beta)$} \;
		\Case{$\alpha \wedge \beta$}{
			create an empty list $r$ \;
			\ForEach{model $m$ in $satisfy(\alpha)$}{
				union $r$ with $satisfy(\beta)$ \;
			}
			return $r$ \;
		}
		\Case{$\alpha \to \beta$}{
			\lIf{$\mathbb{M}_\lambda \models \alpha$}{return $satisfy(\beta)$ \;}
			\lElse{return an empty list \;}
		}
		\Case{$R[\vec x]$}{
			define a new model $\mathbb{N}$ where $|\mathbb{N}| = |\mathbb{M}|$ \;
			add a new member $\omega$ to $|\mathbb{N}|$ \;
			\lForAll{$P_\mathbb{M}$}{$P_\mathbb{N} = P_\mathbb{M}$ \;}
			define $R_{\mathbb{N}}[x_0 \ldots x_n]$ as $R_{\mathbb{M}}[\lambda(x_0) \ldots \lambda(x_n)]$ \;
			\ForEach{$v \in \vec x$}{
				\lIf{$v \not\in \lambda$}{
					$\lambda$ becomes $\lambda_{v \mapsto \omega}$ \;
				}
			}
			return a list containing $\mathbb{N}$ \;
		}
		\Case{$\exists\ \vec x : \alpha$}{
			\lIf{$\vec x = \emptyset$}{ recurse on $\alpha$ } \;
			\eIf{$|\mathbb{M}| \ne \emptyset$ and $\exists\ v' \in |\mathbb{M}| : (\lambda' = \lambda_{x_0 \mapsto v'}$ and $\mathbb{M} \models_{\lambda'} \alpha)$}{
				return a list containing $\mathbb{M}$ \;
			}{
				define a new model $\mathbb{N}$ where $|\mathbb{N}| = |\mathbb{M}|$ \;
				add a member $\omega$ to $|\mathbb{N}|$ such that $\omega \not\in |\mathbb{N}|$ \;
				\lForAll{$R_\mathbb{M}$}{$R_\mathbb{N} = R_\mathbb{M}$ \;}
				define $\kappa = \lambda_{x_0 \mapsto \omega}$ \;
				using model $\mathbb{N}$ and environment $\kappa$, return $satisfy(\exists\ \{x_1 \ldots x_n\} : \alpha)$ \;
			}
		}
	}
	\end{algorithm}

	\subsection{Input Format}

		Input to the parser must be in a form parsable by the following
		context-free grammar. Terminals are denoted by a {\tt monospace style}
		and nonterminals are denoted by an $oblique\ style$. The greek letter
		$\varepsilon$ matches a zero-length list of tokens. Patterns that match
		non-literal terminals are defined in the table following the grammar.

		\textbf{Should these be appendices?}

		\begin{tabular}{ll}
		$program$ & :    $\varepsilon$ \\
		{} & $\mid$      $exprList$ $optNEWLINE$ \\
		\\

		$exprList$ & :   $expr$ \\
		{} & $\mid$      $exprList$ \tt{NEWLINE} $expr$ \\
		\\

		$expr$ & :       \tt{TAUTOLOGY} \\
		{} & $\mid$      \tt{CONTRADICTION} \\
		{} & $\mid$      $expr$ \tt{OR} $expr$ \\
		{} & $\mid$      $expr$ \tt{AND} $expr$ \\
		{} & $\mid$      \tt{NOT} $expr$ \\
		{} & $\mid$      $expr$ \tt{->} $expr$ \\
		{} & $\mid$      \tt{->} $expr$ \\
		{} & $\mid$      $atomic$ \\
		{} & $\mid$      \tt{VARIABLE EQ VARIABLE} \\
		{} & $\mid$      \tt{FOR\_ALL} $argList$ $optCOLON$ $expr$ \\
		{} & $\mid$      \tt{THERE\_EXISTS} $argList$ $optCOLON$ $expr$ \\
		{} & $\mid$      \tt{(} $expr$ \tt{)} \\
		{} & $\mid$      \tt{[} $expr$ \tt{]} \\
		\\

		$atomic$ & :     \tt{PREDICATE} $optIndex$ \\
		\\

		$index$ & :      \tt{(} $argList$ \tt{)} \\
		{} & $\mid$      \tt{[} $argList$ \tt{]} \\
		\\

		$argList$ & :    $arg$ \\
		{} & $\mid$      $argList$ \tt{,} $arg$ \\
		\\

		$arg$ & :        \tt{VARIABLE} \\
		\\

		$optIndex$ & :   $\varepsilon$ \\
		{} & $\mid$      $index$ \\
		\\

		$optCOLON$ & :   $\varepsilon$ \\
		{} & $\mid$      \tt{:} \\
		\\

		$optNEWLINE$ & : $\varepsilon$ \\
		{} & $\mid$      \tt{NEWLINE} \\
		\end{tabular}

		{\tt \begin{tabular}{|l|l|}
			\hline
			\textbf{Input Pattern} & \textbf{Terminal} \\
			\hline
			|                    & OR  \\
			\&                   & AND \\
			!                    & NOT \\
			=                    & EQ  \\
			$[$Tt$]$autology     & TAUTOLOGY \\
			$[$Cc$]$ontradiction & CONTRADICTION \\
			$[\backslash$r$\backslash$n$]$+ & NEWLINE \\
			$[$a-z$][$A-Za-z0-9\_'$]$* & VARIABLE \\
			$[$A-Z$][$A-Za-z0-9\_'$]$* & PREDICATE \\
			For$[$Aa$]$ll        & FOR\_ALL \\
			Exists               & THERE\_EXISTS \\
			\hline
		\end{tabular} }

		Comments are removed at the lexical analysis step and have no effect on
		the input to the parser.  Single-line comments begin with either a hash
		({\tt \#}) or double-dash ({\tt --}). Multiline comments begin with
		{\tt /*} and are terminated by {\tt */}.

	\subsection{Options}

		Help on the usage of the chase implementation can be found by passing
		the executable output by Haskell the {\tt --help} or {\tt -?} options.

		When no options are given to the executable, it expects input from
		stdin and outputs models in a human-readable format to stdout. To take
		input from a file instead, pass the executable the {\tt -i} or {\tt
		--input} option followed by the filename.

		To output models to numbered files in a directory, pass the {\tt -o} or
		{\tt --output} option along with an optional directory name. The given
		directory does not have to exist. If the output directory is omitted,
		it defaults to ``{\tt ./models}".

		Using the {\tt -o} or {\tt --output} options will change the selection
		for output format to a machine-readable format. To switch ouput formats
		at any time, pass the {\tt -h} or {\tt -m} flags for human-readable and
		machine-readable formats respectively.


	\subsection{Future Considerations}

		This section details areas of possible improvement/development.

		\subsubsection{Better Data Structures}

			Some less-than-optimal data structures are being used to hold data
			that should really be in a Data.Map or Data.Set. One such example
			of this is with the truth table holding the relation information of
			a model. This truth table should probably be implemented as a
			Data.Map. Also, instead of Domains being a list of DomainMember, it
			would probably be better if a Domain was a Data.Set.

		\subsubsection{Broader use of the Maybe Monad}

			In several helper functions, the program's execution is halted and
			an error is output when the function receives certain invalid
			inputs. These functions should take advantage of the Maybe Monad
			and return {\tt Maybe a} where they return {\tt a}. One particular
			example of this is the {\tt pef} function. When {\tt pef} takes a
			formula as input that can not be converted to postive-existential
			form, it causes the program to produce an error and exit. Instead,
			{\tt pef} should return {\tt Maybe Formula} and the places where it
			is used should handle the error condition however they choose.

		\subsubsection{Binding Search Approach for Satisfaction Checking}

			When checking if a model with 30 domain members and 60 facts (a
			reasonable real-life usage example) holds for a formula with 5
			universally quantified variables, $30^5$ or $24,300,000$ bindings
			will have to be generated, and the formula will be checked under
			each one. But by limiting the checked bindings to only those that
			can produce facts that exist in the model from the atomics in the
			formula, only the 60 facts will have to be checked in the worst
			case, and only 1 in the best. In practical use, this should
			dramatically reduce the running time of theories that produce
			finite results.

		\subsubsection{Avoid Isomorphic Model Generation}

			By taking different paths to arrive at the same model, the chase
			often creates equivalent models that satisfy its given theory.
			These duplicate models are already being filtered from the output.
			However, another problem exists when the chase returns isomorphic
			models. Given two models, it is very computationally expensive to
			determine if there exists an isomorphism between them. If a fast
			method of determining if two models are isomorphic is found,
			including an implementation of it would provide more valuable
			results to the user.

		\subsubsection{Improve Efficiency}

			The main goals of this project were to write a correct chase
			implementation and apply it in a real-world situation. While
			reasonable and obvious optimizations were made, there is still
			plenty of room for optimization.

			In Harrison's Practical Logic and Automated Reasoning
			\textbf{reference}, Harrison mentions a large number of formula
			rewriting and simplification algorithms, many of which are already
			implemented in the Helpers module. Those functions that are
			written, however, are not currently being used by the chase
			functions, and there are surely other functions that Harrison
			mentions that have not yet been written.

			One specific example of a function that Harrison mentions is
			{\tt pullQuants} \textbf{reference}, which pulls all quantifiers in
			a formula to the outside. This results in a formula with no
			conjunctions or disjunctions of quantified subformul{\ae}. This
			function is implemented, but a function that does exactly the
			opposite of this will help speed up satisfaction checking when
			using the traditional looping method. This will minimize the number
			of times bindings need to be generated for all permutations of the
			current model's domain because the number of quantified variables
			will be reduced.

		\subsubsection{Tracing}

			Currently, Debug.Trace is being used to output real-time status for
			ease of debugging.  This output is helpful to both developers of
			the chase implementation and theories that will be given as input.
			Unfortunately, all of the output can really slow down a chase run
			of a simple theory.  Currently, the only way to disable tracing is
			to replace the definition \[trace = Debug.Trace.trace\] with
			\[trace\ x = id\]

			A flag is already being read in from the command line as {\tt -d}
			and stored in the options record under {\tt optDebug}.
